# Data                                                             :appendix:
:PROPERTIES:
:tangle: R/17-appendix.R
:CUSTOM_ID: cha-data
:END:

This appendix is a reference of sorts regarding some of the data
structures a statistician is likely to encounter. We discuss their
salient features and idiosyncrasies.

## Data Structures
:PROPERTIES:
:CUSTOM_ID: sec-Data-Structures
:END: 

### Vectors

See the "Vectors and Assignment" section of /An Introduction to/
\(\mathsf{R}\). A vector is an ordered sequence of elements, such as
numbers, characters, or logical values, and there may be =NA='s
present. We usually make vectors with the assignment operator =<-=.

```{r, echo=TRUE, results=TRUE} 
x <- c(3, 5, 9)
```

#+RESULTS:

Vectors are atomic in the sense that if you try to mix and match
elements of different modes then all elements will be coerced to the
most convenient common mode.

```{r, echo=TRUE, results=TRUE} 
y <- c(3, "5", TRUE)
```

#+RESULTS:

In the example all elements were coerced to /character/ mode. We can
test whether a given object is a vector with =is.vector= and can
coerce an object (if possible) to a vector with =as.vector=.

### Matrices and Arrays

See the "Arrays and Matrices" section of /An Introduction to/
\(\mathsf{R}\). Loosely speaking, a matrix is a vector that has been
reshaped into rectangular form, and an array is a multidimensional
matrix. Strictly speaking, it is the other way around: an array is a
data vector with a dimension attribute (=dim=), and a matrix is the
special case of an array with only two dimensions. We can construct a
matrix with the =matrix= function.

```{r, echo=TRUE, results=TRUE} 
matrix(letters[1:6], nrow = 2, ncol = 3)
```

#+RESULTS:
:      [,1] [,2] [,3]
: [1,] "a"  "c"  "e" 
: [2,] "b"  "d"  "f"

Notice the order of the matrix entries, which shows how the matrix is
populated by default. We can change this with the =byrow= argument:

```{r, echo=TRUE, results=TRUE} 
matrix(letters[1:6], nrow = 2, ncol = 3, byrow = TRUE)
```

#+RESULTS:
:      [,1] [,2] [,3]
: [1,] "a"  "b"  "c" 
: [2,] "d"  "e"  "f"

We can test whether a given object is a matrix with =is.matrix= and
can coerce an object (if possible) to a matrix with =as.matrix=. As a
final example watch what happens when we mix and match types in the
first argument:

```{r, echo=TRUE, results=TRUE} 
matrix(c(1,"2",NA, FALSE), nrow = 2, ncol = 3)
```

#+RESULTS:
:      [,1] [,2]    [,3]
: [1,] "1"  NA      "1" 
: [2,] "2"  "FALSE" "2"

Notice how all of the entries were coerced to character for the final
result (except =NA=). Also notice how the four values were /recycled/
to fill up the six entries of the matrix.

The standard arithmetic operations work element-wise with matrices.

```{r, echo=TRUE, results=TRUE} 
A <- matrix(1:6, 2, 3)
B <- matrix(2:7, 2, 3)
A + B
A # B
```

#+RESULTS:
:      [,1] [,2] [,3]
: [1,]    3    7   11
: [2,]    5    9   13
:      [,1] [,2] [,3]
: [1,]    2   12   30
: [2,]    6   20   42

If you want the standard definition of matrix multiplication then use
the =%*%= function. If we were to try =A %*% B= we would get an error
because the dimensions do not match correctly, but for fun, we could
transpose =B= to get conformable matrices. The transpose function =t=
only works for matrices (and data frames).

```{r, echo=TRUE, results=TRUE} 
try(A # B)     # an error
A %*% t(B)     # this is alright
```

#+RESULTS:
:      [,1] [,2] [,3]
: [1,]    2   12   30
: [2,]    6   20   42
:      [,1] [,2]
: [1,]   44   53
: [2,]   56   68

To get the ordinary matrix inverse use the =solve= function: 

```{r, echo=TRUE, results=TRUE} 
solve(A %*% t(B))     # input matrix must be square
```

#+RESULTS:
:           [,1]      [,2]
: [1,]  2.833333 -2.208333
: [2,] -2.333333  1.833333

Arrays more general than matrices, and some functions (like transpose)
do not work for the more general array. Here is what an array looks
like:

```{r, echo=TRUE, results=TRUE} 
array(LETTERS[1:24], dim = c(3,4,2))
```

#+RESULTS:
#+BEGIN_example
, , 1

     [,1] [,2] [,3] [,4]
[1,] "A"  "D"  "G"  "J" 
[2,] "B"  "E"  "H"  "K" 
[3,] "C"  "F"  "I"  "L" 

, , 2

     [,1] [,2] [,3] [,4]
[1,] "M"  "P"  "S"  "V" 
[2,] "N"  "Q"  "T"  "W" 
[3,] "O"  "R"  "U"  "X"
#+END_example

We can test with =is.array= and may coerce with =as.array=.

### Data Frames

A data frame is a rectangular array of information with a special
status in \(\mathsf{R}\). It is used as the fundamental data structure
by many of the modeling functions. It is like a matrix in that all of
the columns must be the same length, but it is more general than a
matrix in that columns are allowed to have different modes.

```{r, echo=TRUE, results=TRUE} 
x <- c(1.3, 5.2, 6)
y <- letters[1:3]
z <- c(TRUE, FALSE, TRUE)
A <- data.frame(x, y, z)
A
```

#+RESULTS:
:     x y     z
: 1 1.3 a  TRUE
: 2 5.2 b FALSE
: 3 6.0 c  TRUE

Notice the =names= on the columns of =A=. We can change those with the
=names= function.

```{r, echo=TRUE, results=TRUE} 
names(A) <- c("Fred","Mary","Sue")
A
```

#+RESULTS:
:   Fred Mary   Sue
: 1  1.3    a  TRUE
: 2  5.2    b FALSE
: 3  6.0    c  TRUE

Basic command is =data.frame=. You can test with =is.data.frame= and
you can coerce with =as.data.frame=.

### Lists
A list is more general than a data frame.

### Tables
The word "table" has a special meaning in \(\mathsf{R}\). More
precisely, a contingency table is an object of class =table= which is
an array.

Suppose you have a contingency table and would like to do descriptive
or inferential statistics on it. The default form of the table is
usually inconvenient to use unless we are working with a function
specially tailored for tables. Here is how to transform your data to a
more manageable form, namely, the raw data used to make the table.

First, we coerce the table to a data frame with: 

```{r, echo=TRUE, results=TRUE} 
A <- as.data.frame(Titanic)
head(A)
```

#+RESULTS:
:   Class    Sex   Age Survived Freq
: 1   1st   Male Child       No    0
: 2   2nd   Male Child       No    0
: 3   3rd   Male Child       No   35
: 4  Crew   Male Child       No    0
: 5   1st Female Child       No    0
: 6   2nd Female Child       No    0

Note that there are as many preliminary columns of =A= as there are
dimensions to the table. The rows of =A= contain every possible
combination of levels from each of the dimensions. There is also a
=Freq= column, which shows how many observations there were at that
particular combination of levels.

The form of =A= is often sufficient for our purposes, but more often
we need to do more work: we would usually like to repeat each row of
=A= exactly the number of times shown in the =Freq= column. The
=reshape= package \cite{reshape} has the function =untable= designed
for that very purpose:

```{r, echo=TRUE, results=TRUE} 
B <- with(A, untable(A, Freq))
head(B)
```

#+RESULTS:
:     Class  Sex   Age Survived Freq
: 3     3rd Male Child       No   35
: 3.1   3rd Male Child       No   35
: 3.2   3rd Male Child       No   35
: 3.3   3rd Male Child       No   35
: 3.4   3rd Male Child       No   35
: 3.5   3rd Male Child       No   35

Now, this is more like it. Note that we slipped in a call to the
=with= function, which was done to make the call to =untable= more
pretty; we could just as easily have done
:  untable(TitanicDF, A$Freq)


The only fly in the ointment is the lingering =Freq= column which has
repeated values that do not have any meaning any more. We could just
ignore it, but it would be better to get rid of the meaningless column
so that it does not cause trouble later. While we are at it, we could
clean up the =rownames=, too.

```{r, echo=TRUE, results=TRUE} 
C <- B[, -5]
rownames(C) <- 1:dim(C)[1]
head(C)
```

#+RESULTS:
:   Class  Sex   Age Survived
: 1   3rd Male Child       No
: 2   3rd Male Child       No
: 3   3rd Male Child       No
: 4   3rd Male Child       No
: 5   3rd Male Child       No
: 6   3rd Male Child       No

### More about Tables
Suppose you want to make a table that looks like this:

There are at least two ways to do it.

- Using a matrix:
  ```{r, echo=TRUE, results=TRUE} 
  tab <- matrix(1:6, nrow = 2, ncol = 3)
  rownames(tab) <- c('first', 'second')
  colnames(tab) <- c('A', 'B', 'C')
  tab  # Counts
  ```

  #+RESULTS:
  :        A B C
  : first  1 3 5
  : second 2 4 6

   - note that the columns are filled in consecutively by default. If
     you want to fill the data in by rows then do =byrow = TRUE= in
     the =matrix= command.

- Using a dataframe
  ```{r, echo=TRUE, results=TRUE} 
  p <- c("milk","tea")
  g <- c("milk","tea")
  catgs <- expand.grid(poured = p, guessed = g)
  cnts <- c(3, 1, 1, 3)
  D <- cbind(catgs, count = cnts)
  xtabs(count ~ poured + guessed, data = D)
  ```

  #+RESULTS:
  :       guessed
  : poured milk tea
  :   milk    3   1
  :   tea     1   3

   - again, the data are filled in column-wise.
   - the object is a dataframe
   - if you want to store it as a table then do =A <- xtabs(count ~
     poured + guessed, data = D)=

## Importing Data
:PROPERTIES:
:CUSTOM_ID: sec-Importing-A-Data
:END: 

Statistics is the study of data, so the statistician's first step is
usually to obtain data from somewhere or another and read them into
\(\mathsf{R}\). In this section we describe some of the most common
sources of data and how to get data from those sources into a running
\(\mathsf{R}\) session.

For more information please refer to the \(\mathsf{R}\) /Data
Import/Export Manual/, \cite{rstatenv} and /An Introduction to/
\(\mathsf{R}\), \cite{Venables2010}.

### Data in Packages

There are many data sets stored in the =datasets= package
\cite{datasets} of base \(\mathsf{R}\). To see a list of them all
issue the command =data(package = "datasets")=. The output is omitted
here because the list is so long. The names of the data sets are
listed in the left column. Any data set in that list is already on the
search path by default, which means that a user can use it immediately
without any additional work.

There are many other data sets available in the thousands of
contributed packages. To see the data sets available in those packages
that are currently loaded into memory issue the single command
=data()=. If you would like to see all of the data sets that are
available in all packages that are installed on your computer (but not
necessarily loaded), issue the command

:  data(package = .packages(all.available = TRUE))

To load the data set =foo= in the contributed package =bar= issue the
commands =library("bar")= followed by =data(foo)=, or just the single
command

:  data(foo, package = "bar")

### Text Files
Many sources of data are simple text files. The entries in the file
are separated by delimeters such as TABS (tab-delimeted), commas
(comma separated values, or =.csv=, for short) or even just white
space (no special name). A lot of data on the Internet are stored with
text files, and even if they are not, a person can copy-paste
information from a web page to a text file, save it on the computer,
and read it into \(\mathsf{R}\).

### Other Software Files
Often the data set of interest is stored in some other, proprietary,
format by third-party software such as Minitab, SAS, or SPSS. The
=foreign= package \cite{foreign} supports import/conversion from many
of these formats. Please note, however, that data sets from other
software sometimes have properties with no direct analogue in
\(\mathsf{R}\). In those cases the conversion process may lose some
information which will need to be reentered manually from within
\(\mathsf{R}\). See the /Data Import/Export Manual/.

As an example, suppose the data are stored in the SPSS file =foo.sav=
which the user has copied to the working directory; it can be imported
with the commands

```{r, echo=TRUE, eval=FALSE}
library("foreign")
read.spss("foo.sav")
```

See =?read.spss= for the available options to customize the file
import. Note that the \(\mathsf{R}\) Commander will import many of the
common file types with a menu driven interface.

### Importing a Data Frame

The basic command is =read.table=.

## Creating New Data Sets
:PROPERTIES:
:CUSTOM_ID: sec-Creating-New-Data
:END: 

Using =c=
Using =scan=
Using the \(\mathsf{R}\) Commander.

## Editing Data
:PROPERTIES:
:CUSTOM_ID: sec-Editing-Data-Sets
:END: 

### Editing Data Values
### Inserting Rows and Columns
### Deleting Rows and Columns
### Sorting Data

We can sort a vector with the =sort= function. Normally we have a data
frame of several columns (variables) and many, many rows
(observations). The goal is to shuffle the rows so that they are
ordered by the values of one or more columns. This is done with the
=order= function.

For example, we may sort all of the rows of the =Puromycin= data (in
ascending order) by the variable =conc= with the following:

```{r, echo=TRUE, results=TRUE} 
Tmp <- Puromycin[order(Puromycin$conc), ]
head(Tmp)
```

#+RESULTS:
:    conc rate     state
: 1  0.02   76   treated
: 2  0.02   47   treated
: 13 0.02   67 untreated
: 14 0.02   51 untreated
: 3  0.06   97   treated
: 4  0.06  107   treated

We can accomplish the same thing with the command 

```{r, echo=TRUE, eval=FALSE}
with(Puromycin, Puromycin[order(conc), ])
```

We can sort by more than one variable. To sort first by =state= and
next by =conc= do

```{r, echo=TRUE, eval=FALSE}
with(Puromycin, Puromycin[order(state, conc), ])
```

If we would like to sort a numeric variable in descending order then
we put a minus sign in front of it.

```{r, echo=TRUE, results=TRUE} 
Tmp <- with(Puromycin, Puromycin[order(-conc), ])
head(Tmp)
```

#+RESULTS:
:    conc rate     state
: 11 1.10  207   treated
: 12 1.10  200   treated
: 23 1.10  160 untreated
: 9  0.56  191   treated
: 10 0.56  201   treated
: 21 0.56  144 untreated

If we would like to sort by a character (or factor) in decreasing
order then we can use the =xtfrm= function which produces a numeric
vector in the same order as the character vector.

```{r, echo=TRUE, results=TRUE} 
Tmp <- with(Puromycin, Puromycin[order(-xtfrm(state)), ])
head(Tmp)
```

#+RESULTS:
:    conc rate     state
: 13 0.02   67 untreated
: 14 0.02   51 untreated
: 15 0.06   84 untreated
: 16 0.06   86 untreated
: 17 0.11   98 untreated
: 18 0.11  115 untreated

## Exporting Data
:PROPERTIES:
:CUSTOM_ID: sec-Exporting-a-Data
:END: 

The basic function is =write.table=. The =MASS= package \cite{MASS}
also has a =write.matrix= function.

## Reshaping Data
:PROPERTIES:
:CUSTOM_ID: sec-Reshaping-a-Data
:END: 

- Aggregation
- Convert Tables to data frames and back

=rbind=, =cbind=
=ab[order(ab[,1]),]=
=complete.cases=
=aggregate=
=stack=

