# Writing Reports with \(\mathsf{R}\)                              :appendix:
:PROPERTIES:
:tangle: R/17-appendix.R
:CUSTOM_ID: cha-Writing-Reports-with
:END:

Perhaps the most important part of a statistician's job once the
analysis is complete is to communicate the results to others. This is
usually done with some type of report that is delivered to the client,
manager, or administrator. Other situations that call for reports
include term papers, final projects, thesis work, /etc/. This chapter
is designed to pass along some tips about writing reports once the
work is completed with \(\mathsf{R}\).

## What to Write
:PROPERTIES:
:CUSTOM_ID: sec-What-to-Write
:END: 

It is possible to summarize this entire appendix with only one
sentence: /the statistician's goal is to communicate with others/. To
this end, there are some general guidelines that I give to students
which are based on an outline originally written and shared with me by
Dr. G. Andy Chang.

### Basic Outline for a Statistical Report
1. Executive Summary (a one page description of the study and
   conclusion)
1. Introduction 
   1. What is the question, and why is it important?
   1. Is the study observational or experimental?
   1. What are the hypotheses of interest to the researcher?
   1. What are the types of analyses employed? (one sample \(t\)-test,
      paired-sample \(t\)-test, ANOVA, chi-square test, regression,
      ...)
1. Data Collection 
   1. Describe how the data were collected in detail.
   1. Identify all variable types: quantitative, qualitative, ordered
      or nominal (with levels), discrete, continuous.
   1. Discuss any limitations of the data collection procedure. Look
      carefully for any sources of bias.
1. Summary Information
   1. Give numeric summaries of all variables of interest.
      1. Discrete: (relative) frequencies, contingency tables, odds
         ratios, /etc/.
      1. Continuous: measures of center, spread, shape.
   1. Give visual summaries of all variables of interest.
      1. Side-by-side boxplots, scatterplots, histograms, /etc/.
   1. Discuss any unusual features of the data (outliers, clusters,
      granularity, /etc/.)
   1. Report any missing data and identify any potential problems or
      bias.
1. Analysis 
   1. State any hypotheses employed, and check the assumptions. 
   1. Report test statistics, /p/-values, and confidence intervals. 
   1. Interpret the results in the context of the study.
   1. Attach (labeled) tables and/or graphs and make reference to them
      in the report as needed.
1. Conclusion
   1. Summarize the results of the study. What did you learn?
   1. Discuss any limitations of the study or inferences.
   1. Discuss avenues of future research suggested by the study. 

## How to Write It with R
:PROPERTIES:
:CUSTOM_ID: sec-How-to-Write
:END: 

Once the decision has been made what to write, the next task is to
typeset the information to be shared. To do this the author will need
to select software to use to write the documents. There are many
options available, and choosing one over another is sometimes a matter
of taste. But not all software were created equal, and \(\mathsf{R}\)
plays better with some applications than it does with others.  In
short, \(\mathsf{R}\) does great with \LaTeX{} and there are many
resources available to make writing a document with \(\mathsf{R}\) and
\LaTeX{} easier. But \LaTeX{} is not for the beginner, and there are
other word processors which may be acceptable depending on the
circumstances.

### Microsoft\(\circledR\) Word
It is a fact of life that Microsoft\(\circledR\) Windows is currently
the most prevalent desktop operating system on the planet. Those who
own Windows also typically own some version of Microsoft Office, thus
Microsoft Word is the default word processor for many, many people.

The standard way to write an \(\mathsf{R}\) report with
Microsoft\(\circledR\) Word is to generate material with
\(\mathsf{R}\) and then copy-paste the material at selected places in
a Word document. An advantage to this approach is that Word is nicely
designed to make it easy to copy-and-paste from =RGui= to the Word
document.

A disadvantage to this approach is that the R input/output needs to be
edited manually by the author to make it readable for others. Another
disadvantage is that the approach does not work on all operating
systems (not on Linux, in particular). Yet another disadvantage is
that Microsoft\(\circledR\) Word is proprietary, and as a result,
\(\mathsf{R}\) does not communicate with Microsoft\(\circledR\) Word
as well as it does with other software as we shall soon see.

Nevertheless, if you are going to write a report with Word there are
some steps that you can take to make the report more amenable to the
reader.

1. Copy and paste graphs into the document. You can do this by right
   clicking on the graph and selecting =Copy as bitmap=, or =Copy as
   metafile=, or one of the other options. Then move the cursor to the
   document where you want the picture, right-click, and select
   =Paste=.
1. Resize (most) pictures so that they take up no more than 1/2
   page. You may want to put graphs side by side; do this by inserting
   a table and placing the graphs inside the cells.
1. Copy selected \(\mathsf{R}\) input and output to the Word
   document. All code should be separated from the rest of the
   writing, except when specifically mentioning a function or object
   in a sentence.
1. The font of \(\mathsf{R}\) input/output should be Courier New, or
   some other monowidth font (not Times New Roman or Calibri); the
   default font size of =12= is usually too big for \(\mathsf{R}\)
   code and should be reduced to, for example, =10pt=.

It is also possible to communicate with \(\mathsf{R}\) through
OpenOffice.org, which can export to the proprietary (=.doc=) format.

### OpenOffice.org and \texttt{odfWeave}
OpenOffice.org (OO.o) is an open source desktop productivity suite
which mirrors Microsoft\(\circledR\) Office. It is especially nice
because it works on all operating systems. OO.o can read most document
formats, and in particular, it will read =.doc= files. The standard
OO.o file extension for documents is =.odt=, which stands for "open
document text".

The =odfWeave= package \cite{odfWeave} provides a way to generate an
=.odt= file with \(\mathsf{R}\) input and output code formatted
correctly and inserted in the correct places, without any additional
work. In this way, one does not need to worry about all of the trouble
of typesetting \(\mathsf{R}\) output. Another advantage of =odfWeave=
is that it allows you to generate the report dynamically; if the data
underlying the report change or are updated, then a few clicks (or
commands) will generate a brand new report.

One disadvantage is that the source =.odt= file is not easy to read,
because it is difficult to visually distinguish the noweb parts (where
the \(\mathsf{R}\) code is) from the non-noweb parts. This can be
fixed by manually changing the font of the noweb sections to, for
instance, Courier font, size =10pt=. But it is extra work. It would be
nice if a program would discriminate between the two different
sections and automatically typeset the respective parts in their
correct fonts. This is one of the advantages to \LyX{}.

Another advantage of OO.o is that even after you have generated the
outfile, it is fully editable just like any other =.odt= document. If
there are errors or formatting problems, they can be fixed at any
time.

Here are the basic steps to typeset a statistical report with OO.o.

1. Write your report as an =.odt= document in OO.o just as you would
   any other document. Call this document =infile.odt=, and make sure
   that it is saved in your working directory.
1. At the places you would like to insert \(\mathsf{R}\) code in the
   document, write the code chunks in the following format:
   \texttt{<\textcompwordmark{}<>\textcompwordmark{}>=}~\\
   \texttt{x <- rnorm(10)}~\\
   \texttt{mean(x)}~\\
   \texttt{@} or write whatever code you want between the symbols
   \texttt{<\textcompwordmark{}<>\textcompwordmark{}>=} and
   \texttt{@}.
1. Open \(\mathsf{R}\) and type the following:
   ```{r, echo=TRUE, eval=FALSE}
   library("odfWeave")
   odfWeave(file = "infile.odt", dest = "outfile.odt")
   ```
1. The compiled (=.odt=) file, complete with all of the \(\mathsf{R}\)
   output automatically inserted in the correct places, will now be
   the file =outfile.odt= located in the working directory. Open
   =outfile.odt=, examine it, modify it, and repeat if desired.

There are all sorts of extra things that you can do. For example, the
\(\mathsf{R}\) commands can be suppressed with the tag
\texttt{<\textcompwordmark{}<echo = FALSE>\textcompwordmark{}>=}, and
the \(\mathsf{R}\) output may be hidden with
\texttt{<\textcompwordmark{}<results =
hide>\textcompwordmark{}>=}. See the =odfWeave= package documentation
for details.

### Sweave and \protect\LaTeX{}

This approach is nice because it works for all operating systems. One
can quite literally typeset /anything/ with \LaTeX{}. All of this
power comes at a price, however. The writer must learn the \LaTeX{}
language which is a nontrivial enterprise. Even given the language, if
there is a single syntax error, or a single delimeter missing in the
entire document, then the whole thing breaks.

\LaTeX{} can do anything, but it is relatively difficult to learn and
very grumpy about syntax errors and delimiter matching. There are
however programs useful for formatting LaTeX.

A disadvantage is that you cannot see the mathematical formulas until
you run the whole file with \LaTeX{}.

A disadvantage is that figures and tables are relatively difficult.

There are programs to make the process easier: AUC\TeX{}.

dev.copy2eps, also dev.copy2pdf

[[http://www.stat.uni-muenchen.de/~leisch/Sweave/]]

### Sweave and LyX
This approach is nice because it works for all operating systems. It
gives you everything from the last section and makes it easier to use
\LaTeX{}. That being said, it is better to know \LaTeX{} already when
migrating to \LyX{}, because you understand all of the machinery going
on under the hood.

Program Listings and the \(\mathsf{R}\) language
[[http://gregor.gorjanc.googlepages.com/lyx-sweave]]

## Formatting Tables
:PROPERTIES:
:CUSTOM_ID: sec-Formatting-Tables
:END: 

The prettyR package \cite{prettyR}.

The Hmisc package \cite{Hmisc}.

```{r, echo=TRUE, results=TRUE} 
summary(cbind(Sepal.Length, Sepal.Width) ~ Species, data = iris)
```

#+RESULTS:
#+BEGIN_example
cbind(Sepal.Length, Sepal.Width)    N=150

+-------+----------+---+------------+-----------+
|       |          |N  |Sepal.Length|Sepal.Width|
+-------+----------+---+------------+-----------+
|Species|setosa    | 50|5.006000    |3.428000   |
|       |versicolor| 50|5.936000    |2.770000   |
|       |virginica | 50|6.588000    |2.974000   |
+-------+----------+---+------------+-----------+
|Overall|          |150|5.843333    |3.057333   |
+-------+----------+---+------------+-----------+
#+END_example

There is a =method= argument to =summary=, which is set to
\texttt{method = "response"} by default. There are two other methods
for summarizing data: =reverse= and =cross=. See =?summary.formula= or
the [[http://biostat.mc.vanderbilt.edu/twiki/bin/view/Main/StatReport][following document]] from Frank Harrell for more details.

## Other Formats
:PROPERTIES:
:CUSTOM_ID: sec-Other-Formats
:END: 

HTML and prettyR
R2HTML
